Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-08-05 paper/manuscript.tex
approaching final v0.1

\todototoc
\listoftodos

%%%%%%%%%%%%%%%%%%%%
%  Introduction    %
%%%%%%%%%%%%%%%%%%%%
The current Overleaf document is \url{https://www.overleaf.com/5136551543vgjqbkzpmqkp}
\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities. 
 These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein sidechain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.

@@

%%%%%%%%%%%%%%%%%%%%
% Prerequesites    %
%%%%%%%%%%%%%%%%%%%%
\section{Prerequisites and Scope}
\label{sec:pre}
This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this propose we provide a convenient checklist (Section~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
 It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.

@@

As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Section~\ref{sec:theory}.
In the remainder of this paper, we will cover topics that are key to the preparation( Section~\ref{sec:prerequisites}), choice and use of correct protocols (Section~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Section~\ref{sec:data_analysis}). 
Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).

@@

\subsection{Simulating binding events of receptor-drug systems can be computationally expensive}
Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a receptor $R$, given by:
\begin{equation}
R+L \leftrightharpoons RL.
\end{equation}
The binding constant ($K_b^{\circ}$) is given by the law of mass action as the ratio of concentrations of product $[RL]$ and reactants $[R]$, $[L]$:
\begin{equation}
 K_b^{\circ} = c^{\circ}\frac{[RL]}{[L][R]}.
\end{equation}
The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Section~\ref{sec:standardstate-restraints}).
Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
\begin{equation}
    \Delta G_{\mathrm{bind},L} = -k_BT\ln K_b^{\circ},
\end{equation}
where $k_B$ is the Boltzmann constant and $T$ the temperature of the system.

@@

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the millisecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

@@

In practice, the mutation of $A$ to $B$ is carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Section~\ref{sec:relative-fe-protocol} for details).

We can rigorously account for fluctuations in other thermodynamic parameters such as changes in volume $V$ when simulating at constant pressure $p$ or changes in number of molecules $N_i$ of species $i$ at constant chemical potential $\mu_i$ (e.g., number of waters or ions) by introducing the \textit{reduced potential}~\cite{shirts2008statisticallya}
\begin{equation}\label{eq:reduced-potential}
u(\vec{q};\vec{\lambda}) \equiv \beta \left[ U(\vec{q};\vec{\lambda}) + p \, V(\vec{q}) + \sum_i \mu_i \, N_i(\vec{q}) + \cdots \right] \, .
\end{equation}
Here, the collection of thermodynamic and alchemical parameters $\{\beta, \vec{\lambda}, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
In the context of alchemical calculations, in which the thermodynamic states vary only in their value of $\vec{\lambda}$, these are also referred to as \emph{alchemical states}.
The free energy of mutating $A$ to $B$ in any environment (e.g., binding site, solvent) can then be computed as
\begin{equation}\label{eq:delta-G-env}
    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, .
\end{equation}
While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Section~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.

\paragraph{Why do alchemical calculations need unphysical intermediate states?}
While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state (see Section~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).

To solve the problem of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
Each intermediate state models a ligand that is neither $A$ nor $B$ but a mix of the two.
Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennet's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
\begin{equation}
    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1})
\end{equation}
from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
\begin{equation}
    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
\end{equation}
While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates. This is achieved by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Section~\ref{subsec:estimators}).

@@

\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of example datasets and case studies can be found in Section~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

It is worth noting that the accuracy of free energy calculations is highly variable across different protein targets, and likely across different ligand chemotypes as well.
For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, but an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}.
Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}. Here the errors refer to the difference in $\Delta G$ after doing a linear regression fit to observed experimental binding free energies. This prompts us to consider another important aspect. It is important to be clear on what error to report: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges. (See recommendations for reporting best practices, Sec.~\ref{sec:plot_data}) In general, the computed edge $\Delta \Delta G$ RMSEs significantly underestimates the error of all edge RMSEs: \url{https://github.com/jchodera/jacs-dataset-analysis}. Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each application case.

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
Finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol)~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. A good practice is to perform two or three runs of the same perturbation to assess reproducibility with a given protocol. The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat run.

@@

For illustrative purposes, assume that the actual changes in the
binding free energies for a set of new molecules also follows a normal
distribution, that the standard deviation in the changes is $RT\ln 5$
(corresponding to a 5-fold change in the binding affinities), and that
1 in 10 new molecules have increased binding affinity ($\Delta\Delta G
\leq 0$). Under such assumptions, the conditional probability in
Eq.~\ref{eqn:true-positive} can be easily computed.

@@

If you are unsure whether your project is feasible, as mentioned above, one recommended option is to conduct a short exploratory study to assess tractability for a small
number of ligands. This can be sufficient to get an initial
idea of feasibility and accuracy of the calculations for the
proposed target.

@@

\subsection{Capturing experimental conditions}
The calculations aim to capture the alchemical change from one ligand to another as accurately as possible. Therefore, it is necessary to consider details of the experimental setup, such as pH. Biological assays are usually run at neutral pH but this is not always the case. For example, some enzymes exhibit pH-dependent activity and assays may thus be done in conditions other from neutral pH. Therefore, computational protein and ligand preparation protocols should reflect experimental pH.

@@

It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions. A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}. Beyond this, several parameters provide useful indicators of performance. For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}. Hysteresis, within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}). Once synthesis and testing of compounds is complete it would be normal to look back at how the calculations performed. In this regard it is important to consider upfront the issue of selection bias. It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{abel2017critical}. For a more detailed discussion on checking the robustness of your alchemcial free energy calculation see also Section~\ref{sec:are-they-good}.

@@

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. 
From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult. In the meantime, the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical calculations.

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Section~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of binding or a free energy of hydration. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

\paragraph{Are there good X-ray structures available?}
As with any simulation, care should be taken in selecting available X-ray structures in the Protein DataBank. In some cases it may be wise to choose multiple starting structures to account for variability in receptor conformations as well as the accuracy of available X-ray structures. Typically, clustering of receptor structures can be used to identify different receptor conformations near the binding site, as well as assessing relevant side chain placements from the X-ray structure, see for example~\cite{mey2016blinded}. In terms of set up and other choices, following general best practice guidelines is advisable~\cite{braun2019best}.

Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see Section~\ref{sec:simulation_protocol_choice}). For relative calculations, some care has to be taken selecting binding poses for these ligands. Generally, a common assumption for a congeneric series is that the binding mode is conserved. Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation without steric or electrostatic mismatch with the receptor. Checking the X-ray structure versus the experimental electron densities is important, as the position of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density, especially in cases of missing density. For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and, if a boat conformation is present in the structure, it may be worth inspecting the density to ensure it adequately supports this choice.

>>>>

%\todototoc
%\listoftodos

%%%%%%%%%%%%%%%%%%%%
%  Introduction    %
%%%%%%%%%%%%%%%%%%%%

\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or the effects of protein side chain mutations on binding affinities or thermostabilities. 
 These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligandreceptor}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or protein sidechain) is changed by modifying the potential governing the interactions with the environment for the atoms being modified, inserted, or deleted.

@@

%%%%%%%%%%%%%%%%%%%%
% Prerequesites    %
%%%%%%%%%%%%%%%%%%%%
\section{Prerequisites and Scope}
\label{sec:pre}
This Best Practices guide focuses on providing a good starting point for new practitioners and a reference for experienced practitioners. 
 For this propose we provide a convenient checklist (Sec.~\ref{sec:checklist}) to help ensure all calculations comply with currently-understood best practices for alchemical simulation and analysis. Where the best practices are currently not certain, we highlight areas where further research is needed to identify an unambiguous recommendation.
 It can also serve as a set of best practices to ensure simulation robustness and reproducibility which reviewers may wish to consider as they evaluate papers.

@@

As some of the theoretical background can seem daunting, we do, however, provide a guide to the essential theory behind alchemical free energy calculations in Sec.~\ref{sec:theory}.
In the remainder of this paper, we will cover topics that are key to the preparation( Sec.~\ref{sec:prerequisites}), choice and use of correct protocols (Sec.~\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used in the analysis of alchemical calculations (Sec.~\ref{sec:data_analysis}). 
Particular focus will be given to aspects of the molecular simulations which are unique to alchemical calculations---these include the calculation of transfer free energies (hydration free energies, partition coefficients, etc.), and binding free energies (absolute and relative).

@@

\subsection{Simulating binding events of receptor-drug systems can be computationally expensive}
Suppose you want to compute the binding affinity, or free energy of binding, of a ligand $L$ to a receptor $R$, given by:
\begin{equation}
R+L \leftrightharpoons RL.
\end{equation}
The binding constant ($K_b^{\circ}$) is given by the law of mass action as the ratio of concentrations of product $[RL]$ and reactants $[R]$, $[L]$:
\begin{equation}
 K_b^{\circ} = c^{\circ}\frac{[RL]}{[L][R]}.
\end{equation}
The standard state concentration $c^\circ$ depends on the reference state, but it is usually set to $1\,\mathrm{mol}/\mathrm{L}$ assuming a constant pressure of $1\,\mathrm{atm}$ (see also Sec.~\ref{sec:standardstate-restraints}).
Thus, the Gibbs free energy of binding $\Delta G_{\mathrm{bind}}$ is given by:
\begin{equation}
    \Delta G_{\mathrm{bind},L} = -k_BT\ln K_b^{\circ},
\end{equation}
where $k_B$ is the Boltzmann constant and $T$ the temperature of the system.

@@

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the micosecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

@@

In practice, the mutation of $A$ to $B$ is carried out by introducing one or more parameters $\vec{\lambda}$ controlling the potential energy function $U(\vec{q};\vec{\lambda})$ such that the potential of compounds $A$ and $B$ is recovered at two particular values $\vec{\lambda}_A$ and $\vec{\lambda}_B$.
Briefly, this is achieved by simulating a ``chimeric'' molecule composed of enough atoms to represent both $A$ and $B$.
A subset of the energetic terms in $U(\vec{q};\vec{\lambda})$ is then modulated by $\vec{\lambda}$ so that at $\vec{\lambda}_A$, the atoms that form molecule $A$ are activated and those belonging exclusively to $B$ are non-interacting ``dummy atoms'', while the opposite occur at $\vec{\lambda}_B$ (see Sec.~\ref{sec:relative-fe-protocol} for details).

We can rigorously account for fluctuations in other thermodynamic parameters such as changes in volume $V$ when simulating at constant pressure $p$ or changes in number of molecules $N_i$ of species $i$ at constant chemical potential $\mu_i$ (e.g., number of waters or ions) by introducing the \textit{reduced potential}~\cite{shirts2008statisticallya}
\begin{equation}\label{eq:reduced-potential}
u(\vec{q};\vec{\lambda}) \equiv \beta \left[ U(\vec{q};\vec{\lambda}) + p \, V(\vec{q}) + \sum_i \mu_i \, N_i(\vec{q}) + \cdots \right] \, .
\end{equation}
Here, the collection of thermodynamic and alchemical parameters $\{\beta, \vec{\lambda}, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
In the context of alchemical calculations, in which the thermodynamic states vary only in their value of $\vec{\lambda}$, these are also referred to as \emph{alchemical states}.
The free energy of mutating $A$ to $B$ in any environment (e.g., binding site, solvent) can then be computed as
\begin{equation}\label{eq:delta-G-env}
    \Delta G_{\mathrm{env}} = - k_BT \ln \frac{Z(\vec{\lambda}_B)}{Z(\vec{\lambda}_A)} = - k_BT \ln \frac{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_B) \right) \, d\vec{q}}{\int_{\Gamma_{\mathrm{env}}} \exp\left( u(\vec{q}; \vec{\lambda}_A) \right) \, d\vec{q}} \, .
\end{equation}
While it is generally not feasible to compute the two partition functions $Z(\vec{\lambda})$, several estimators have been devised to robustly estimate the ratio of partition functions in Eq.~\ref{eq:delta-G-env} (see Sec.~\ref{subsec:estimators}) from a set of conformations usually collected with MD simulations from the thermodynamic states defined at $\vec{\lambda}_A$ and $\vec{\lambda}_B$ and intermediates thereof.

\paragraph{Why do alchemical calculations need unphysical intermediate states?}
While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability conformations (i.e., those with very negative potential energy) in one state are also high-probability conformations in the other state (see Sec.~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).

To solve the problem of having poor overlap between the states of interest, multiple intermediate alchemical states are introduced at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
Each intermediate state models a ligand that is neither $A$ nor $B$ but a mix of the two.
Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennet's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
\begin{equation}
    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1})
\end{equation}
from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
\begin{equation}
    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
\end{equation}
While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya} can exploit similarities between states to improve the precision of the estimates. This is achieved by using the conformations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).

@@

\subsection{How accurate are alchemical free energy calculations?}
\label{subsec:expectation}
Current alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of example datasets and case studies can be found in Sec.~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

It is worth noting that the accuracy of free energy calculations is highly variable across different protein targets, and likely across different ligand chemotypes as well.
For instance, FEP+ with OPLS3 achieves an RMSE of 0.62 kcal/mol for a set of 21 compounds binding to JNK1 kinase, but an RMSE of 1.05 kcal/mol for a set of 34 compounds binding to P38$\alpha$ kinase~\cite{harder2016opls3}.
Furthermore, perturbations for the same chemotype in different pockets of the BACE enzyme gave varied errors~\cite{keranen2017acylguanidine}. Here the errors refer to the difference in $\Delta G$ derived by fitting the $\Delta \Delta G$'s to the known experimental binding free energies~\cite{wang2015accurate}. This prompts us to consider another important aspect. It is important to be clear on what error to report: $\Delta G$ after shifting by a constant to minimize the RMSE, unshifted $\Delta G$, $\Delta \Delta G$ of computed edges, or $\Delta \Delta G$ of all edges. (See recommendations for reporting best practices, Sec.~\ref{sec:plot_data}.) Additionally, as it is possible to perform calculations on a set of ligands using different pairwise comparisons of molecules, the performance of the method may be biased based on which pairs of comparisons are performed. Additionally, it is possible that the error associated to the relative free energy between a two ligands that was not directly computed (however can be deduced using thermodynamic paths involving other ligands) will likely be more uncertain \url{https://github.com/jchodera/jacs-dataset-analysis}.
Given the need to understand the performance of the system with alchemical free energy calculations, we recommend that retrospective studies for a particular target and a particular chemical series be performed for each application case.

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
Finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package (with a sample standard deviation under 0.01 kcal/mol)~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. A good practice is to perform two or three runs of the same perturbation to assess precision with a given protocol. The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat run.

@@

For illustrative purposes, consider a proposed set of new molecules, and assume that the changes proposed in these molecules yield a set of relative binding free energies that follow a normal
distribution. That is, assume that the standard deviation in the relative binding free enrgies for the changes represented is $RT\ln 5$
(corresponding to a 5-fold change in the binding affinities), and that
1 in 10 new molecules have increased binding affinity ($\Delta\Delta G
\leq 0$). Under such assumptions, the conditional probability in
Eq.~\ref{eqn:true-positive} can be easily computed.

@@

If you are unsure whether your project is feasible, as mentioned above, one recommended option is to conduct a short exploratory study to assess tractability for a small
number of ligands. This can be sufficient to get an initial
idea of feasibility and accuracy of the calculations for the
proposed target~\cite{schindler2020largescale}.

@@

\subsection{Capturing experimental conditions}
\label{subsec:exp_condition}
The calculations aim to capture the alchemical change from one ligand to another as accurately as possible. Therefore, it is necessary to consider details of the experimental setup, such as pH. Biological assays are usually run at neutral pH but this is not always the case. For example, some enzymes exhibit pH-dependent activity and assays may thus be done in conditions other from neutral pH. Therefore, computational protein and ligand preparation protocols should reflect experimental pH.

@@

It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions. A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}. Beyond this, several parameters provide useful indicators of performance. For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}. Hysteresis, within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}). Once synthesis and testing of compounds is complete a standard strategy is to look back at how the calculations performed. In this regard it is important to consider the issue of selection bias upfront. It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{abel2017critical}. For a more detailed discussion on checking the robustness of your alchemcial free energy calculation see also Sec.~\ref{sec:are-they-good}.

@@

Structure based drug design projects are often capable of improving potency relatively quickly, even with only limited application of computational approaches and the range of activity narrows to just two-to-three log units. It may seem hard to have impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has shown the value of activity prediction, see discussion above and articles such as~\cite{mobley2012perspective}. 
From a drug discovery point of view, alchemical calculations are expanding their domain of applicability, and there are reports of success using homology models~\cite{cappel2016relative} and GPCRs~\cite{deflorian2020accurate,lenselink2016predicting} for instance, as well as enabling charge change and scaffold hopping~\cite{chen2018accurate, wang2017accurate}, but these systems are undoubtedly more difficult. In the meantime, the use cases are expanding to resistance prediction, selectivity prediction , solubility prediction – an exciting future for alchemical calculations~\cite{hauser2018predicting, albanese2020structure, mondal2019free}.

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of binding or a free energy of hydration. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

\paragraph{Are there good X-ray structures available?}
As with any simulation, care should be taken in selecting available X-ray structures in the Protein DataBank~\cite{berman2003announcing}. In some cases it may be wise to choose multiple starting structures to account for variability in receptor conformations as well as the accuracy of available X-ray structures. Typically, clustering of receptor structures can be used to identify different receptor conformations near the binding site, as well as assessing relevant side chain placements from the X-ray structure, see for example~\cite{mey2016blinded}. In terms of set up and other choices, following general best practice guidelines is advisable~\cite{braun2019best}.

Many free energy calculations focus on a congeneric series of ligands, which can make these calculations suitable for relative free energy protocols (see Sec.~\ref{sec:simulation_protocol_choice}). For relative calculations, some care has to be taken selecting binding poses for these ligands. Generally, a common assumption for a congeneric series is that the binding mode is conserved. Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation without steric or electrostatic mismatch with the receptor. Checking the X-ray structure versus the experimental electron densities is important, as the position of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density, especially in cases of missing density. For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and, if a boat conformation is present in the structure, it may be worth inspecting the density to ensure it adequately supports this choice.
